### 2018-07-10 submission/manuscript.Rmd
update discussion

Our observations support that the number of OTUs, error rate, and chimera prevalence depends on polymerase used and that chimera prevalence is highly correlated with the number of OTUs [Figure 1-2 & 4-6]. Our observations also suggest that undetected chimeras, not removed using standard bioinformatic approaches, cause many of these differences. In addition to the reported polymerase dependent differences, the bacterial community also varied by the number of cycles used. These differences are important because many diversity metrics rely on calculations (e.g. richness) that are dependent on the number of OTUs. Since the number of OTUs are dependent on error rate and chimera prevalence the diversity metrics can vary simply by changing the polymerase used. Based on these observations, metrics that measure diversity using calculations such as richness depend on the polymerase, but this may not be the case for all diversity metrics.

Based on polymerase used, our observations generally found no difference when using the Bray-Curtis index to measure the bacterial community. Specifically, there was no difference in distance between successive 5-cycle increments within samples between the polymerases when using the Bray-Curtis index [Figure 3]. However, the distance between cycle comparison groups (e.g. 20 cycles versus 25 cycles group compared to 30 cycles versus 35 cycles group) was different across polymerase used. These results support the idea that cycle number and not the specific polymerase used may have a larger affect on metrics such as Bray-Curtis. One possible reason for this outcome is that many of the OTUs generated by polymerase dependent error rates and chimera prevalence are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight these OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate some of the observed polymerase dependent differences in error rate and chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996], richness) may be less robust to polymerase dependent error rates and chimera prevalence. When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account. One possible way to better choose polymerases may be based on the DNA polymerase family used in the PCR mixture.

Although the variation in error rate and chimera prevalence may be due to the DNA polymerase family because of the different binding affinities and error correction capacity [@polymerase_Ishino_2014], this is unlikely to be the only contributor. Within our study the highest and lowest chimera prevalence both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014]. Additionally, based on the information supplied by the respective manufacturers, the differences between the two PCR mixtures are not immediately apparent. Both PCR mixtures contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody. Since it is not possible to know everything about the mixture beyond what was willingly provided by the manufacturer, it is possible that differences in how the recombinant *Taq* was generated or other compounds within the PCR mixture could be a contributing factor for the differences in error rate and chimera prevalence. Beyond the choice of the type of polymerase, there may be other ways to reduce the affect of polymerase dependent error rates and chimera prevalence on the downstream results.

Standardization of the 16S rRNA gene sequencing workflow may partially solve this problem by introducing a consistent bias to all samples. However, as mentioned in the introduction, many of the choices made to standardize the workflow will result in missing important members of the microbial community. Therefore, the degree of workflow standardization across studies and research groups need to be approached on a study by study basis. All aspects of the 16S rRNA gene sequencing workflow should be customized for the specific scientific questions that want to be answered. Although a diversity of approaches may make reproducibility and replicability more difficult it will help reduce systematic biases from occurring.

@@

Our observations show that errors dependent on either high fidelity DNA polymerases used or the number of cycles chosen can change diversity metrics used in 16S rRNA gene sequencing. Care should be taken when choosing a polymerase for 16S rRNA gene surveys because they can influence diversity-based metrics. By knowing the inherent bias associated with different polymerases it allows for more informed choices to be made on how to reduce 16S rRNA gene sequencing-based error.

>>>>

Previous studies using regular DNA polymerase have reported GC dependent amplification differences [@polz_bias_1998] and we confirm that these differences still persist in high fidelity DNA polymerase when analyzing mock community samples [Figure 1]. Additionally, these specific polymerases also have different sequence error rate and chimera prevalence in these samples [Figure 2]. Chimera prevalence in particular was strongly associated with the total number of OTUs detected in a sample [Figure 2E]. The total number of OTUs also varied by polymerase used for mock community samples [Fiugre 3A]. The range in the number of OTUs observed increased with the number of cycles and the change in range was dependent on the polymerase for mock community and fecal samples [Figure 3]. However, these differences did not translate to large bacterial community composition changes based on polymerase [Figure 4]. Even though there were bacterial community composition changes based on the number of cycles they occured independently of the polymerase used [Figure 4]. Additionally, there were no differences observed within the number of cycles between any of the polymerases for fecal samples. When Random Forest models were created to classify number of cycles, polymerase, and individual based on the 16S rRNA gene sequencing the model performance for individuals was significantly greater than all other models tested [Figure 5A]. In addition to the better model performance of individuals the most important OTUs to this model were not the most important OTUs in the other models [Figure 5B]. Our observations show that differences between high fidelity DNA polymerases can be found. However, when using 16S rRNA gene sequencing these technical differences are dwarfed by biological differences between individuals.

The small differences observed due to error rate and chimera prevalence may be due to the actual DNA polymerase family being used within the PCR mix. DNA polymerases from distinct families have different binding affinities and error correction capacity [@polymerase_Ishino_2014]. However, based on the observations within our study this is unlikely to be the only contributor. Within our study the highest and lowest chimera prevalence both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014]. Additionally, based on the information supplied by the respective manufacturers, the differences between the two PCR mixtures are not immediately apparent. Both PCR mixtures contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody. Since it is not possible to know everything about the mixture beyond what was willingly provided by the manufacturer, it is possible that differences in how the recombinant *Taq* was generated or other compounds within the PCR mixture could be a contributing factor for the differences in error rate and chimera prevalence. Beyond the choice of the type of polymerase, there may be other ways to reduce the affect of polymerase dependent error rates and chimera prevalence on the downstream results.

Based on polymerase used, our observations generally found no difference when using the Bray-Curtis index to measure the bacterial community. Specifically, there was no difference in distance between successive 5-cycle increments within samples between the polymerases when using the Bray-Curtis index [Figure 4]. One possible reason for this outcome is that many of the OTUs generated by polymerase dependent error rates and chimera prevalence are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight these OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate some of the observed polymerase dependent differences in error rate and chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996], richness) may be less robust to polymerase dependent error rates and chimera prevalence. When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account. One possible way to better choose polymerases may be based on the DNA polymerase family used in the PCR mixture.

Collectively, our data shows that using distinct high fidelity DNA polymerases will result in a different number of OTUs being detected. The differences in the number of OTUs are primarily due to chimera prevalence and sequence error rates that are distinct to the specific polymerase. Despite this variation in the number of OTUs, our Random Forest models were able to correctly classify individuals regardless of the number of cycles and polymerase used. These results suggest that even though there is a large variation in the high fidelity DNA polymerases used across studies, biological variation similar in effect to bacterial community composition differences between individuals will still be consistent.

@@

Although care should always be taken when choosing a polymerase for 16S rRNA gene sequencing, our observations show that the differences between a variety of polymerases are dwarfed by the actual biological variation between individuals. Therefore, if the biological signal of interest is similar to differences in bacterial community composition found between individuals, then the type of high fidelity DNA polymerase used will only minimally change the results.

### 2018-07-10 submission/manuscript.tex
update discussion

Our observations support that the number of OTUs, error rate, and
chimera prevalence depends on polymerase used and that chimera
prevalence is highly correlated with the number of OTUs {[}Figure 1-2 \&
4-6{]}. Our observations also suggest that undetected chimeras, not
removed using standard bioinformatic approaches, cause many of these
differences. In addition to the reported polymerase dependent
differences, the bacterial community also varied by the number of cycles
used. These differences are important because many diversity metrics
rely on calculations (e.g.~richness) that are dependent on the number of
OTUs. Since the number of OTUs are dependent on error rate and chimera
prevalence the diversity metrics can vary simply by changing the
polymerase used. Based on these observations, metrics that measure
diversity using calculations such as richness depend on the polymerase,
but this may not be the case for all diversity metrics.

Based on polymerase used, our observations generally found no difference
when using the Bray-Curtis index to measure the bacterial community.
Specifically, there was no difference in distance between successive
5-cycle increments within samples between the polymerases when using the
Bray-Curtis index {[}Figure 3{]}. However, the distance between cycle
comparison groups (e.g.~20 cycles versus 25 cycles group compared to 30
cycles versus 35 cycles group) was different across polymerase used.
These results support the idea that cycle number and not the specific
polymerase used may have a larger affect on metrics such as Bray-Curtis.
One possible reason for this outcome is that many of the OTUs generated
by polymerase dependent error rates and chimera prevalence are likely
not highly abundant, allowing the Bray-Curtis index to be able to
successfully down-weight these OTUs (Minchin, 1987). The choice of
downstream diversity metric could be an important consideration in
helping to mitigate some of the observed polymerase dependent
differences in error rate and chimera prevalence. Metrics that solely
use presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas, 1996),
richness) may be less robust to polymerase dependent error rates and
chimera prevalence. When choosing a distance metric, careful
consideration of the biases introduced from the PCR step of the 16S rRNA
gene sequencing workflow need to be taken into account. One possible way
to better choose polymerases may be based on the DNA polymerase family
used in the PCR mixture.

Although the variation in error rate and chimera prevalence may be due
to the DNA polymerase family because of the different binding affinities
and error correction capacity (Ishino \& Ishino, 2014), this is unlikely
to be the only contributor. Within our study the highest and lowest
chimera prevalence both belonged to a family A polymerase (Platinum and
Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
Additionally, based on the information supplied by the respective
manufacturers, the differences between the two PCR mixtures are not
immediately apparent. Both PCR mixtures contain a recombinant \emph{Taq}
DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a platinum
\emph{Taq} antibody. Since it is not possible to know everything about
the mixture beyond what was willingly provided by the manufacturer, it
is possible that differences in how the recombinant \emph{Taq} was
generated or other compounds within the PCR mixture could be a
contributing factor for the differences in error rate and chimera
prevalence. Beyond the choice of the type of polymerase, there may be
other ways to reduce the affect of polymerase dependent error rates and
chimera prevalence on the downstream results.

Standardization of the 16S rRNA gene sequencing workflow may partially
solve this problem by introducing a consistent bias to all samples.
However, as mentioned in the introduction, many of the choices made to
standardize the workflow will result in missing important members of the
microbial community. Therefore, the degree of workflow standardization
across studies and research groups need to be approached on a study by
study basis. All aspects of the 16S rRNA gene sequencing workflow should
be customized for the specific scientific questions that want to be
answered. Although a diversity of approaches may make reproducibility
and replicability more difficult it will help reduce systematic biases
from occurring.

>>>>

Previous studies using regular DNA polymerase have reported GC dependent
amplification differences (Polz \& Cavanaugh, 1998) and we confirm that
these differences still persist in high fidelity DNA polymerase when
analyzing mock community samples {[}Figure 1{]}. Additionally, these
specific polymerases also have different sequence error rate and chimera
prevalence in these samples {[}Figure 2{]}. Chimera prevalence in
particular was strongly associated with the total number of OTUs
detected in a sample {[}Figure 2E{]}. The total number of OTUs also
varied by polymerase used for mock community samples {[}Fiugre 3A{]}.
The range in the number of OTUs observed increased with the number of
cycles and the change in range was dependent on the polymerase for mock
community and fecal samples {[}Figure 3{]}. However, these differences
did not translate to large bacterial community composition changes based
on polymerase {[}Figure 4{]}. Even though there were bacterial community
composition changes based on the number of cycles they occured
independently of the polymerase used {[}Figure 4{]}. Additionally, there
were no differences observed within the number of cycles between any of
the polymerases for fecal samples. When Random Forest models were
created to classify number of cycles, polymerase, and individual based
on the 16S rRNA gene sequencing the model performance for individuals
was significantly greater than all other models tested {[}Figure 5A{]}.
In addition to the better model performance of individuals the most
important OTUs to this model were not the most important OTUs in the
other models {[}Figure 5B{]}. Our observations show that differences
between high fidelity DNA polymerases can be found. However, when using
16S rRNA gene sequencing these technical differences are dwarfed by
biological differences between individuals.

The small differences observed due to error rate and chimera prevalence
may be due to the actual DNA polymerase family being used within the PCR
mix. DNA polymerases from distinct families have different binding
affinities and error correction capacity (Ishino \& Ishino, 2014).
However, based on the observations within our study this is unlikely to
be the only contributor. Within our study the highest and lowest chimera
prevalence both belonged to a family A polymerase (Platinum and
Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
Additionally, based on the information supplied by the respective
manufacturers, the differences between the two PCR mixtures are not
immediately apparent. Both PCR mixtures contain a recombinant \emph{Taq}
DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a platinum
\emph{Taq} antibody. Since it is not possible to know everything about
the mixture beyond what was willingly provided by the manufacturer, it
is possible that differences in how the recombinant \emph{Taq} was
generated or other compounds within the PCR mixture could be a
contributing factor for the differences in error rate and chimera
prevalence. Beyond the choice of the type of polymerase, there may be
other ways to reduce the affect of polymerase dependent error rates and
chimera prevalence on the downstream results.

Based on polymerase used, our observations generally found no difference
when using the Bray-Curtis index to measure the bacterial community.
Specifically, there was no difference in distance between successive
5-cycle increments within samples between the polymerases when using the
Bray-Curtis index {[}Figure 4{]}. One possible reason for this outcome
is that many of the OTUs generated by polymerase dependent error rates
and chimera prevalence are likely not highly abundant, allowing the
Bray-Curtis index to be able to successfully down-weight these OTUs
(Minchin, 1987). The choice of downstream diversity metric could be an
important consideration in helping to mitigate some of the observed
polymerase dependent differences in error rate and chimera prevalence.
Metrics that solely use presence/absence of OTUs (e.g.~Jaccard (Real \&
Vargas, 1996), richness) may be less robust to polymerase dependent
error rates and chimera prevalence. When choosing a distance metric,
careful consideration of the biases introduced from the PCR step of the
16S rRNA gene sequencing workflow need to be taken into account. One
possible way to better choose polymerases may be based on the DNA
polymerase family used in the PCR mixture.

Collectively, our data shows that using distinct high fidelity DNA
polymerases will result in a different number of OTUs being detected.
The differences in the number of OTUs are primarily due to chimera
prevalence and sequence error rates that are distinct to the specific
polymerase. Despite this variation in the number of OTUs, our Random
Forest models were able to correctly classify individuals regardless of
the number of cycles and polymerase used. These results suggest that
even though there is a large variation in the high fidelity DNA
polymerases used across studies, biological variation similar in effect
to bacterial community composition differences between individuals will
still be consistent.
